\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-28 1-intro.Rmd
proofread abstract adn intro

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time interval. This type of data can naturally been considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

While the data is enriched the information in both the spatial and temporal dimension, it also brings variables in two different observational units, which poses the problem of organising data for various operations in exploratory data analysis, visualisation, and modelling. The single table format is convenient for spatio-temporal operations, however, it makes duplicates of spatial variables and is not convenient for spatial operations. The two tables format allows separate spatial and temporal operations on each table while analysts need to work with two data objects simultaneously and dedicate efforts to make sure the sites in both tables are matched up. This data organising issue can also raise during the analysis, for example, a spatial summary of the temporal or spatio-temporal data naturally introduces an additional spatial table and makes the data in the separate table format. With the current tools, at each step, analysts need to decide on the format to operate on, perform the operation, and organise the result with existing data.

@@

In the \proglang{R} community, different data objects have been proposed to work with spatio-temporal data when the specific task is known, while few have the capability to freely explore the data for exploratory data analysis. \pkg{spacetime} [@spacetime] is built on class \pkg{sp} [@sp] and \pkg{xts} [@xts] and both of which have been replaced by more recent spatial and temporal data objects. A more implementation, \pkg{stars} [@stars], uses a spatio-temporal array for wrangling both vector and raster data, but for raw data already in tabular format, need to convert the data into the array structure, also convert at vis. While there has not yet been a data object that makes it easy to explore spatio-temporal data that overcomes the difficulties of wrangling spatio-temporal data, this motivates developing a new tool that hopefully can solve this problem. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild: climate observations from weather stations, air quality variables from air monitoring stations, or any data collected at fixed location across time. The requirement can also comes from spatial data analysts wanting to incorporate a temporal dimension in their data or vice versa.

This paper proposes an \proglang{R} package, \pkg{cubble}, for exploratory data analysis in spatio-temporal data. The proposed new data structure, also called \code{cubble}, overcomes the inconvenience in data wrangling mentioned above through wiring the separate spatial and temporal table into a single data object. With the new data object, analysts can spend less time organising spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

>>>>

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time intervals. This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variable. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

@@

While incorporating both spatial and temporal information enriches the data in the analysis, spatio-temporal data can sometimes be hard to work with given its two observational units in the variables. The single table format in Figure \ref{fig:illu-input} is convenient for spatio-temporal operations that involve both types of variables, however, it makes duplicates of spatial variables and is not ideal when only looking at the spatial aspect of the data. The two tables format allows separate operations on spatial and temporal variables while analysts need to work with two data objects simultaneously and join them together into the single table format for spatio-temporal operations. Neither of the two formats is desirable at the same time for spatial, temporal, and spatio-temporal operations. This issue implies analysts need to constantly think about the data format and arrange variables into the proper shape before the actual operations, which can be a laborious step when switching between the two formats back and forth several times during the analysis.

@@

In the \proglang{R} community, many data objects have been proposed to work with specific spatio-temporal tasks, while few can freely explore the data. \pkg{spacetime} [@spacetime] is an attempt for such a data object, but the spatial and temporal class it built on have now been gradually replaced by more recent implementations. The \pkg{stars} [@stars] package uses an array to wrangle spatio-temporal data, but the dense array it uses can quickly exhaust the memory for data in the irregular spatial grid. While there has not yet been a data object that makes it easy to wrangle spatio-temporal data, there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to overcome the problem with organising variables in different levels when working with spatio-temporal data. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With the new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

### 2022-02-28 main.Rmd
proofread abstract adn intro

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refers to measurements taken across space over time. In practise, spatial and temporal data many times need to be combined in order to form a spatio-temporal data set. However, these two types of data are collected at different observational units and this poses a problem on which format to use for different data wrangling tasks. Analysts can either organised the data in a single table format that combines all the variables together, or in two separate tables, each focusing on the spatial or temporal aspect of the data. Both formats are useful in exploratory data analysis, visualisation, and, modelling, however it can be labourious to constantly switch between the two formats during the analysis and write code to cast the data into the desireable shape. This paper presents a new \proglang{R} package, \pkg{cubble}, that provides a new data sturcture to ease the transformation between the two formats in spatio-temporal data. The new data structure organises spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while syncronised as a whole piece. \pkg{Cubble} also provides capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics and performing spatio-temporal transformation. The paper will also demonstrate \pkg{cubble} with examples using Australia climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

>>>>

---
documentclass: jss
author:
  - name: H. Sherry Zhang
    affiliation: Monash University
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email: \email{huize.zhang@monash.edu}
  - name: Dianne Cook
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    # location: Melbourne, Australia
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: 'University of Natural Resources and Life Sciences \AND'
    address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
    email:  ursula.laa@boku.ac.at
  - name: Nicolas Langrené
    affiliation: 'BNU-HKBU United International College \AND'
    address: 2000 Jintong Road, Tangjiawan, Zhuhai, Guangdong Province, China
    email: nicolas@langrene.com
  - name: Patricia Menéndez
    affiliation: 'Monash University \AND'
    address: 21 Chancellors Walk, Clayton VIC 3800 Australia
    email:  patricia.menendez@monash.edu
title:
  formatted: "\\pkg{cubble}: An R Package for Structuring Spatio-temporal Data"
  # If you use tex in the formatted title, alsoge supply version without
  plain:     "cubble: An R Package for Structuring Spatio-temporal Data"
  # For running headers, if needed
  # short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  Spatio-temporal data refers to measurements taken across space over time. In practice, spatial and temporal data many times need to be combined in order to form a spatio-temporal data set. However, these two types of data are collected at different observational units and this poses a problem on how to structure the data for different wrangling tasks. Analysts can either organise the data in a single table format that combines all the variables, or in two separate tables, each focusing on the spatial or temporal aspect of the data. Both formats are useful in exploratory data analysis, visualisation, and, modelling, however, it can be labourious to constantly switch between the two formats during the analysis and write code to cast the data into the desirable shape. This paper presents a new \proglang{R} package, \pkg{cubble}, that provides a new data structure to ease the conversion between the two formats in spatio-temporal data. The new data structure organises spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. \pkg{Cubble} also provides the capability for handling data with hierarchical structure, matching data from multiple sources, constructing interactive graphics, and performing spatio-temporal transformation. The paper will demonstrate \pkg{cubble} with examples using Australian climate weather stations, river level data, and climate reanalysis (ERA5) data.
keywords:
  # at least one keyword must be supplied
  formatted: [spatial, temporal, spatio temporal, "\\proglang{R}", environmental data, exploratory data analysis]
  plain:     [spatial, temporal, spatio temporal, R, environmental data, exploratory data analysis]
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
  \usepackage{xcolor}
output: rticles::jss_article
bibliography: references.bib
---

### 2022-02-28 main.tex
proofread abstract adn intro

\Abstract{
Spatio-temporal data refers to measurements taken across space over
time. In practise, spatial and temporal data many times need to be
combined in order to form a spatio-temporal data set. However, these two
types of data are collected at different observational units and this
poses a problem on which format to use for different data wrangling
tasks. Analysts can either organised the data in a single table format
that combines all the variables together, or in two separate tables,
each focusing on the spatial or temporal aspect of the data. Both
formats are useful in exploratory data analysis, visualisation, and,
modelling, however it can be labourious to constantly switch between the
two formats during the analysis and write code to cast the data into the
desireable shape. This paper presents a new \proglang{R} package,
\pkg{cubble}, that provides a new data sturcture to ease the
transformation between the two formats in spatio-temporal data. The new
data structure organises spatial and temporal variables as two forms of
a single data object so that they can be wrangled separately while
syncronised as a whole piece. \pkg{Cubble} also provides capability for
handling data with hierarchical structure, matching data from multiple
sources, constructing interactive graphics and performing
spatio-temporal transformation. The paper will also demonstrate
\pkg{cubble} with examples using Australia climate weather stations,
river level data, and climate reanalysis (ERA5) data.
}

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time interval. This type of data can naturally
been considered as a data cube with three axes: spatial identifier,
temporal identifier, and variable. In this cubic formation, space only
occupies one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

While the data is enriched the information in both the spatial and
temporal dimension, it also brings variables in two different
observational units, which poses the problem of organising data for
various operations in exploratory data analysis, visualisation, and
modelling. The single table format is convenient for spatio-temporal
operations, however, it makes duplicates of spatial variables and is not
convenient for spatial operations. The two tables format allows separate
spatial and temporal operations on each table while analysts need to
work with two data objects simultaneously and dedicate efforts to make
sure the sites in both tables are matched up. This data organising issue
can also raise during the analysis, for example, a spatial summary of
the temporal or spatio-temporal data naturally introduces an additional
spatial table and makes the data in the separate table format. With the
current tools, at each step, analysts need to decide on the format to
operate on, perform the operation, and organise the result with existing
data.

In the \proglang{R} community, different data objects have been proposed
to work with spatio-temporal data when the specific task is known, while
few have the capability to freely explore the data for exploratory data
analysis. \pkg{spacetime} \citep{spacetime} is built on class \pkg{sp}
\citep{sp} and \pkg{xts} \citep{xts} and both of which have been
replaced by more recent spatial and temporal data objects. A more
implementation, \pkg{stars} \citep{stars}, uses a spatio-temporal array
for wrangling both vector and raster data, but for raw data already in
tabular format, need to convert the data into the array structure, also
convert at vis. While there has not yet been a data object that makes it
easy to explore spatio-temporal data that overcomes the difficulties of
wrangling spatio-temporal data, this motivates developing a new tool
that hopefully can solve this problem. The requirement for such a tool
is important given the ubiquity of spatio-temporal vector data in the
wild: climate observations from weather stations, air quality variables
from air monitoring stations, or any data collected at fixed location
across time. The requirement can also comes from spatial data analysts
wanting to incorporate a temporal dimension in their data or vice versa.

This paper proposes an \proglang{R} package, \pkg{cubble}, for
exploratory data analysis in spatio-temporal data. The proposed new data
structure, also called \code{cubble}, overcomes the inconvenience in
data wrangling mentioned above through wiring the separate spatial and
temporal table into a single data object. With the new data object,
analysts can spend less time organising spatial and temporal variables
and focus more on the exploratory data analysis itself. The software is
available from the Comprehensive R Archive Network (CRAN) at {[}CRAN
link{]}.

>>>>

\Abstract{
Spatio-temporal data refers to measurements taken across space over
time. In practice, spatial and temporal data many times need to be
combined in order to form a spatio-temporal data set. However, these two
types of data are collected at different observational units and this
poses a problem on how to structure the data for different wrangling
tasks. Analysts can either organise the data in a single table format
that combines all the variables, or in two separate tables, each
focusing on the spatial or temporal aspect of the data. Both formats are
useful in exploratory data analysis, visualisation, and, modelling,
however, it can be labourious to constantly switch between the two
formats during the analysis and write code to cast the data into the
desirable shape. This paper presents a new \proglang{R} package,
\pkg{cubble}, that provides a new data structure to ease the conversion
between the two formats in spatio-temporal data. The new data structure
organises spatial and temporal variables as two forms of a single data
object so that they can be wrangled separately while synchronised as a
whole piece. \pkg{Cubble} also provides the capability for handling data
with hierarchical structure, matching data from multiple sources,
constructing interactive graphics, and performing spatio-temporal
transformation. The paper will demonstrate \pkg{cubble} with examples
using Australian climate weather stations, river level data, and climate
reanalysis (ERA5) data.
}

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variable. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

@@

While incorporating both spatial and temporal information enriches the
data in the analysis, spatio-temporal data can sometimes be hard to work
with given its two observational units in the variables. The single
table format in Figure \ref{fig:illu-input} is convenient for
spatio-temporal operations that involve both types of variables,
however, it makes duplicates of spatial variables and is not ideal when
only looking at the spatial aspect of the data. The two tables format
allows separate operations on spatial and temporal variables while
analysts need to work with two data objects simultaneously and join them
together into the single table format for spatio-temporal operations.
Neither of the two formats is desirable at the same time for spatial,
temporal, and spatio-temporal operations. This issue implies analysts
need to constantly think about the data format and arrange variables
into the proper shape before the actual operations, which can be a
laborious step when switching between the two formats back and forth
several times during the analysis.

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal class it built on have now
been gradually replaced by more recent implementations. The \pkg{stars}
\citep{stars} package uses an array to wrangle spatio-temporal data, but
the dense array it uses can quickly exhaust the memory for data in the
irregular spatial grid. While there has not yet been a data object that
makes it easy to wrangle spatio-temporal data, there is a strong need
for such tools given the ubiquitous spatial and temporal data collected
in the society, i.e., climate observations from weather stations, air
quality variables from air monitoring stations. The demand for such
tools can also come from spatial data analysts wishing to incorporate
temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With the new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.
